Binning Strategies and Related Loss for Large Data

### 2013-04-03 Paper.Rnw
edits to binning and discussion sections

Mathematically, binning corresponds to a function $b(.)$ that assigns to a value $x$ the closest bin center $x^\ast \in \{ x_j^\ast \}, j \in \{1, ..., n_X\}$, i.e. the binning function   $b(.) : x \rightarrow x^\ast$ is defined as
\begin{eqnarray*}\label{stdbin}
x^\ast = b(x) =  b_X \cdot \left\lfloor (x-O_X)/ b_X + b_X/2\right\rfloor + O_X
\end{eqnarray*}
where $O_X$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. 
%to allow for bins of width $b_X$ centered at any sequence $O_X, O_X + b_X, ..., O_X + n b_X$.

For illustration, let us assume that we have observed values 1.23, 1.55, and 2.35, and bin centers $\{1, 2, 3\}$, corresponding to a bin width of $b_X = 1$ and a first bin center $O_X = 1$.
Intuitively it is clear, that value 1.23 is assigned to bin center 1, while values 1.55 and 2.35 are assigned to bin center 2.
Applying equation (\ref{stdbin}) to value 1.23 we get: $1 \cdot \lfloor (1.23-1)/ 1 + 0.5\rfloor + 1$ = $\lfloor 0.73 \rfloor + 1$ =  0 + 1 = 1. Similarly, 1.55 is assigned to $\lfloor 0.55 + 0.5\rfloor + 1$ = 1+1 = 2.

Alternate binning parameterizations include dividing up the range of data by a specified number of bins of equal width, but even these strategies fit into the framework of assigning values to the closest bin center.
% , with smallest bin
% $b_X   \cdot \lfloor (x_{(1)}-O_X)/b_X + 0.5\rfloor + O_X$ and largest bin 
% $b_X   \cdot \lfloor (x_{(n)}-O_X)/b_X + 0.5\rfloor + O_X$.

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo,echo=F,include=F,cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

The method described above is the standard binning algorithm. As a different approach, we define the random binning algorithm below.

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Bin Attribute Diagram]{\includegraphics[keepaspectratio=TRUE, width=4in]
  {./figure/Binning-DataVisualNumericalCenter.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

@@

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@
An interesting variation which results in statistically nice properties is a random, non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x$ to bin center $x^\ast$ with a probability depending on the distance to that bin center;  the closer a value is to a bin center, the higher the probability that the value is assigned to this bin center. 
More formally,
\begin{eqnarray}\label{randbin}
b^r(x) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x)}/b_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x - x^\ast_{j})}/b_X
  \end{array}\right.
\end{eqnarray}

@@

Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.

@@

The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary. The standard algorithm is preferable because it has lower total combined numerical and visual loss, but the two algorithms are very similar when we compare frequency loss. The random algorithm does have more variability, but the frequency loss values are much more similar, and when bins are of size two in either or both dimensions, the random algorithm has lower frequency loss than the standard algorithm. This holds for bins which are powers of two, as long as the random binning algorithm is applied iteratively to take advantage of the constant loss property noted above. %~\ref{fig:LossComparison}

@@

>>>>

Mathematically, binning corresponds to a function $b(.)$ that assigns to a value $x$ the closest bin center $x^\ast \in \{ x_j^\ast \}, j \in \{1, ..., n_X\}$, i.e. the binning function   $b(.) : x \rightarrow x^\ast$ is defined as
\begin{eqnarray*}\label{stdbin}
x^\ast = b(x) =  b_X \cdot \left\lfloor (x-x_1^\ast)/ b_X + 1/2\right\rfloor + x_1^\ast
\end{eqnarray*}
where $x_{(1)}$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. An intuitive understanding of the formula is developed by noting that $(x-x_1^\ast)/ b_X$ is the number of bin widths that $x$ is above $x_1^\ast$ and that $\frac{b_X/2}{b_X}=\frac{1}{2}$ is the number of bin widths that $x_1^\ast$ is above $x_{(1)}$. The value $\lfloor (x-x_1^\ast)/ b_X + 1/2\rfloor$ then represents the number of bins centers that lie between $x_{(1)}$ and $x$.  Thus, the binning function is simply finding the location of the bin center for $x$ by adding the distance between the bin centers for $x_{(1)}$ and $x$ to the known bin center for $x_{(1)}$.
%to allow for bins of width $b_X$ centered at any sequence $O_X, O_X + b_X, ..., O_X + n b_X$.

Alternate binning parameterizations include dividing up the range of data by a specified number of bins of equal width, but even these strategies fit into the framework of assigning values to the closest bin center.

% , with smallest bin
% $b_X   \cdot \lfloor (x_{(1)}-O_X)/b_X + 0.5\rfloor + O_X$ and largest bin 
% $b_X   \cdot \lfloor (x_{(n)}-O_X)/b_X + 0.5\rfloor + O_X$.

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process extended to a two dimensional situation.
<<binningDemo,echo=F,include=F,cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Bin Attribute Diagram]{\includegraphics[keepaspectratio=TRUE, width=4in]
  {./figure/Binning-DataVisualNumericalCenter.pdf}}
  \caption{In illustration (a) points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region. Illustration (b) displays the points assigned to a single bin along with key bin attributes: the visual bin center, ($x_1^\ast$, $y_1^\ast$), and the numerical bin center, ($\bar{x}$,$\bar{y}$). }
  \label{fig:BinningTutorial}
\end{figure}

@@

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@
An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties. A random binning algorithm utilizes a non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x$ to bin center $x^\ast$ with a probability depending on the distance to that bin center;  the closer a value is to a bin center, the higher the probability that the value is assigned to this bin center. 
More formally,
\begin{eqnarray}\label{randbin}
b^r(x) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x)}/b_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x - x^\ast_{j})}/b_X
  \end{array}\right.
\end{eqnarray}

@@

%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.

@@

Frequency loss is argueably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary. The standard and random algorithms are very similar when we compare frequency loss. The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticable than the increase seen in numerical loss. %~\ref{fig:LossComparison}

@@

THIS SECTION UNDER CONSTRUCTION
